From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-07-30 WOLFHPC15/article/concept.tex
concepts fix helene 30_07

%----------------------------------------
\subsection{Stencil codes}
\label{sect:stencil}
%----------------------------------------
To numerically solve a set of PDEs, iterative methods (finite difference, finite volume, finite element methods etc.) are frequently used to approximate the solution through a discretized (step by step) phenomena. Thus, the continuous time and space domains are discretized so that a set of numerical computations are iteratively (time discretization) applied onto a mesh (space discretization). In other words, in a mesh-based numerical simulation, the PDEs are transformed to a set of numerical computations applied at each time step on elements of the discretized space domain (the mesh). Among those numerical computations is found a set of numerical schemes, also called \textit{stencil computations}~\cite{spaaTangCKLL11}. To ease programming of stencil computations, and because of the relative regularity of such a computation, many solutions (languages or libraries) propose an easy way to describe the stencil code, while producing transparently an optimized and parallelized stencil code. For example, Liszt~\cite{DeVito2011LDS} and OP2~\cite{Giles2011} are domain specific languages to define a stencil code on unstructured meshes, while Pochoir~\cite{spaaTangCKLL11} and PATUS~\cite{citeulike12258902} address structured meshes, respectively proposing an optimized cache tiling technique, and a separation of concerns by separating parallelization strategy from the computation description.

Let introduce some concepts used in this paper. First, a stencil code is based on a \emph{mesh} which is the discretization of the physical real domain. A mesh is a connected undirected graph without bridges (an edge is a bridge if its removal results in two disconnected graphs), where nodes and edges are linked to form cells (closure). For example, a structured mesh is illustrated in Figure~\ref{fig:mesh}, each cell contains four nodes and four edges. Second, \emph{mesh domains} are subset elements of the mesh, such as the center of cells, edges or nodes, for example. In Figure~\ref{fig:mesh} two mesh domains are illustrated, the center of the cells (in red) and edges on the horizontal axe (in blue on left and right of each cell). A \emph{data} is a quantity to simulate or a temporary data to perform computations. A data is mapped onto a given mesh domain. For example, in Figure~\ref{fig:ex1}, data $A$ and $B$ are mapped onto the center of cells, while in Figure~\ref{fig:ex2} $C$ is mapped onto the edges.

@@

Finally, a stencil computation is composed of a \emph{numerical expression} which indicates how to compute an element of the computation domain (on the output data), using elements of the set of input data. The numerical expression of a stencil computation has the particularity to use not only the element to compute but also a \emph{neighborhood} of this element, also called a \emph{stencil shape}. For example, the stencil shape of the first computation in Figure~\ref{fig:ex1} contains direct neighbors on the right, left, top and bottom. Sometimes, the neighborhood shape can also access a different mesh domain, as for example in Figure~\ref{fig:ex2}. Actually, in this computation, the neighborhood contains edges on the left and on the right of the cell. As an example, the numerical expression of the first example could be
\begin{equation*} 
A(x,y) = B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1),
\end{equation*}
and the numerical expression of the second example could be
\begin{equation*} 
A(x,y) = A(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}

@@

For $\Delta$ the set of data of the simulation, a stencil computation $s$ is defined as the quintuplet
\begin{equation} 
s(R,w,exp,d_c),
\label{eq:st}
\end{equation}
where $R$ is a list of pair $(r,n)$ with $r \in \Delta$ is a data read by the computation, and $n$ is a stencil shape (a neighborhood). The data written by the computation is denoted $w \in \Delta$, $exp$ is the numerical expression of the stencil computation. Finally the numerical expression is applied on the computation domain $d_c$.

>>>>

%----------------------------------------
\subsection{Stencil codes}
\label{sect:stencil}
%----------------------------------------
To numerically solve a set of PDEs, iterative methods (finite difference, finite volume, finite element methods etc.) are frequently used to approximate the solution through a discretized (step by step) phenomena. Thus, the continuous time and space domains are discretized so that a set of numerical computations are iteratively (time discretization) applied onto a mesh (space discretization). In other words, in a mesh-based numerical simulation, the PDEs are transformed to a set of numerical computations applied at each time step on elements of the discretized space domain (the mesh). Among those numerical computations is found a set of numerical schemes, also called \textit{stencil computations}~\cite{spaaTangCKLL11}. To ease programming of stencil computations, and because of the relative regularity of such a computation, many solutions (languages or libraries) propose an easy way to describe the stencil code, while producing transparently an optimized and parallelized stencil code. For example, Liszt~\cite{DeVito2011LDS} and OP2~\cite{Giles2011} are domain specific languages to define a stencil code on unstructured meshes, while Pochoir~\cite{spaaTangCKLL11} and PATUS~\cite{citeulike12258902} address structured meshes, respectively proposing an optimized cache tiling technique, and a separation of concerns between parallelization strategies and computation descriptions.

Let introduce some concepts used in this paper. First, a stencil code is based on a \emph{mesh} which is the discretization of the physical real domain. A mesh is a connected undirected graph without bridges (an edge is a bridge if its removal results in two disconnected graphs), where nodes and edges are linked to form cells (closure). For example, a structured mesh is illustrated in Figure~\ref{fig:mesh}. Each cell contains four nodes and four edges. Second, \emph{mesh domains} are subset elements of the mesh, such as the center of cells, edges or nodes. In Figure~\ref{fig:mesh} two mesh domains are illustrated, the center of the cells (in red) and edges on the horizontal axe (in blue on left and right of each cell). A \emph{data} is a quantity to simulate or a temporary data to perform computations. A data is mapped onto a given mesh domain. For example, in Figure~\ref{fig:ex1}, data $A$ and $B$ are mapped onto the center of cells, while in Figure~\ref{fig:ex2} $C$ is mapped onto the edges.

@@

Finally, a stencil computation is composed of a \emph{numerical expression} which indicates how to compute an element of the computation domain (on the output data), using elements of the set of input data. The numerical expression of a stencil computation has the particularity to use not only the element to compute but also a \emph{neighborhood} of this element, also called a \emph{stencil shape}. For example, the stencil shape of the first computation in Figure~\ref{fig:ex1} contains direct neighbors on the right, left, top and bottom. Sometimes, the neighborhood shape can also access a different mesh domain, as for example in Figure~\ref{fig:ex2}. Actually, in this computation, the neighborhood contains edges on the left and on the right of a cell. As an example, the numerical expression of the first example could be
\begin{equation*} 
A(x,y) = B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1),
\end{equation*}
and the numerical expression of the second example could be
\begin{equation*} 
A(x,y) = A(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}

@@

For $\Delta$ the set of data of the simulation, a stencil computation $s$ is defined as the quintuplet
\begin{equation} 
s(R,w,exp,d_c),
\label{eq:st}
\end{equation}
where $R$ is a set of pair $(r,n)$ with $r \in \Delta$ is a data read by the computation, and $n$ is a stencil shape (a neighborhood) applied on $r$. The data written by the computation is denoted $w \in \Delta$, $exp$ is the numerical expression of the stencil computation. Finally the numerical expression is applied on the computation domain $d_c$.
